Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2018-02-24 HRUIonGPU/2nd submit/jrnl_article_bak.tex
Adding HRUIonGPU article

For a three dimensional beamspace we show that processing rates supporting real-time cardiac ultrasound imaging are possible, meaning that images can be processed faster than the image acquisition rate for a wide range of parameters. Image quality is investigated in an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element-space and beamspace.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

% denne må utelates i IUS-artikkelen
In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes, Buskenes2013} and cardiac ultrasound imaging \cite{asen2012}. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. In \cite{asen2012} and in this paper we show that Capon beamforming applied in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a cluster of high-end GPUs can provide. Since the execution time of Capon beamforming is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we extend the work done in \cite{asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation. The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of element-space Capon when applied in narrow-beam systems.
%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information on Capon beamforming in both element-space and beamspace. We also give an introduction to the GPU computing model. In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}. Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}. Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

@@

As shown in Fig.\ \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. As with coarse-fine beamformers \cite{Thomenius}, the complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

@@

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig.\ \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig.\ \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
%We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space. 
The overlap between subarrays that were exploited when estimating the covariance matrix in element space, can now be utilized when calculating the beamspace transform. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
\begin{align}\label{eq:bs_formula}
x_{\text{BS},l,k} &= \mat{B}_k\vec{x}_l \,\,\,\,\,\,\,\,\, k \in [0, N_b-1] \\
&= \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L},
\end{align}
where $\mat{B}_k$ is the $k$th row of $\mat{B}$. %Manipulating this expression we find that for $l \in [1, L-1]$
%\begin{align}\label{eq:sliding_bs}
%x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
%&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
%\end{align}

%We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
This formula has been implemented using a normalized sliding DFT (SDFT)\cite{Lyons2003}.
%Equation (\ref{eq:sliding_bs}) shows that 
The SDFT allows us to still save instructions by utilizing the overlap between subarrays, but with an increase in task level granularity. In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations. One solution, as depicted in Fig.\ \ref{fig:gpulayoutbs}, is to combine SDFT with full computations. A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SDFT algorithm. In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.

@@

To summarize, by transforming data to beamspace the complexity of all subsequent steps is potentially reduced. Calculation of the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculation of weights and the beamformer output are both $O(N_b)$. The added complexity of SDFT is $O(N_bL)$. To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.

@@

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
\hfil
\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
\label{fig:benchBS96}}}
\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.\ \ref{fig:bench}. When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively. The large jumps in execution times are caused by adaptive kernel launch configurations, which depends on $M$, $L$ and $N_b$.}
\label{fig:benchBS}
\end{figure*}
In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency. Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line. The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007}, 
\begin{equation}\label{eq:rayleigh}
\delta{}\theta = \lambda/(2Mp),
\end{equation}
where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
To sample a $70\degree$ sector with $64$- and $96$-element arrays with $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.

The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory. This is a valid assumption, since the imaging scenarios presented in this paper are well below the 8.0 GB/s limit of PCI Express 2.0. What is included are all calculations, memory transfers from global to shared GPU memory, and shared memory access. The delay step has previously been shown to take far less time than what we report for both Capon beamformers \cite{Song2012, Chen2011}.

Benchmarks of the ES-Capon beamformer are presented in Fig.\ \ref{fig:bench}. For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$. For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 8.3 frames per second (FPS). For $M=96$, $L=48$ and $K=2$ the processing throughput drops below $0.5$ FPS.

In Fig.\ \ref{fig:benchBS} we present similar benchmarks for the BS-Capon beamformer. The figure only presents results with use of temporal averaging ($K=2$). For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 FPS (8 ms). For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 FPS (13 ms). In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using one receive line per transmit. Hence, below this line we are capable of performing real-time processing.

\section{In-vivo cardiac images}\label{sec:images}

%\begin{figure*}[!t]
%\centerline{
%\subfloat[]{ \includegraphics[width=2.2in]{gfx/das.png} \label{fig:phantomDAS} }
%\hfil
%\subfloat[]{ \includegraphics[width=2.2in]{gfx/capon.png} \label{fig:phantomMV} }
%\hfil
%\subfloat[]{ 
%\includegraphics[width=2.2in]{gfx/capon_bs.png} \label{fig:phantomBS} 
%\hfil
%\includegraphics[width=0.35in]{gfx/colorBarPhantom2.png}
%}
%}
%\caption{Simulated phantom containing one bright (20 dB) and one dark (-20 dB) circular area and two point targets (30 dB) in speckle (0 dB). The simulation was done using Field II with a 64-element, 3.4 MHz, and $\lambda/2$ pitch phased array. Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB. The image is the 15th frame in the attached videos. a) Delay-and-sum with uniform weights. b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 1}. c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 2}.}
%\label{fig:phantom}
%\end{figure*}

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=2.2in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
\hfil
\subfloat[]{\includegraphics[width=2.2in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
\hfil
\subfloat[]{
\includegraphics[width=2.2in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
\hfil
\includegraphics[width=0.4in]{gfx/colorBarInvivo2.png}
}
\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5 MHz, 64 elements phased array. Number of receive beams for the $70\degree$ sector is 98, and number of samples in range is 610. Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB. The image is the first frame in the attached videos. a) Delay-and-sum with uniform weights \multimedia{Media-Movie 1}. b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 2}. c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 3}.}
\label{fig:invivo}
\end{figure*}

%\begin{figure*}[!t]
%\centerline{
%\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
%\hfil
%\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
%}
%\caption{Lateral slices taken from the images in Fig. \ref{fig:phantom} and Fig. \ref{fig:invivo}. a) From Fig. \ref{fig:phantom} at 93 mm constant depth, through the two bright point scatterers. b) From Fig. \ref{fig:invivo} at 96 mm range, through the tip of the mitral valve.}
%\label{fig:slices}
%\end{figure*}

%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig. \ref{fig:phantom}). The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth. Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth. The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon. The presented image is frame $15$ in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.

Fig.\ \ref{fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle. The image was acquired using a 64-element and $p=0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive). The image frame is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 1, 2 and 3}.
The region of interest is 70\degree{} with a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}).
The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.  %It is included to show that both ES-Capon and BS-Capon beamforming are feasible for \textit{in-vivo} cardiac imaging, not just in simulations.

%To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig. \ref{fig:phantom} and \ref{fig:invivo} are included in Fig. \ref{fig:slices}. For the phantom, the cut goes through the two point targets at 90 mm constant radius, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant radius.

%Both data sets have 
The data set has been processed with the following settings for the ES-Capon beamformer: $L = 32$, $K = 2$, and $d=0.1$. These settings have previously been shown to provide improved lateral resolution while maintaining DAS-like speckle \cite{Synnevag2007a}. For BS-Capon the following parameters have been used: $L=32$, $K=2$, $d=0.01$, and $N_b=3$.

>>>>

For a three dimensional beamspace we show that processing rates supporting real-time cardiac ultrasound imaging are possible, meaning that images can be processed faster than the image acquisition rate for a wide range of parameters. Image quality is investigated in videos of both a simulated phantom, and in an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element-space and beamspace.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

% denne må utelates i IUS-artikkelen
In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes, Buskenes2013} and cardiac ultrasound imaging \cite{asen2012}. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. In \cite{asen2012} and in this paper we show that Capon beamforming applied in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a cluster of high-end GPUs can provide. Since the execution time of Capon beamforming is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we extend the work done in \cite{asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation. The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of element-space Capon applied in narrow-beam systems.
%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information on Capon beamforming in both element-space and beamspace and the GPU computing model. In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}. Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}. Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

@@

As shown in Fig. \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. As with coarse-fine beamformers \cite{Thomenius}, the complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

@@

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
%We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space. 
The overlap between subarrays that were exploited when estimating the covariance matrix in element space, can now be utilized when calculating the beamspace transform. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
\begin{align}\label{eq:bs_formula}
x_{\text{BS},l,k} &= \mat{B}_k\vec{x}_l \,\,\,\,\,\,\,\,\, k \in [0, N_b-1] \\
&= \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L},
\end{align}
where $\mat{B}_k$ is the $k$th row of $\mat{B}$. Manipulating this expression we find that for $l \in [1, L-1]$
\begin{align}\label{eq:sliding_bs}
x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
\end{align}

We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).

Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity. In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations. One solution, as depicted in Fig. \ref{fig:gpulayoutbs}, is to combine SBS with full computations. A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SBS algorithm. In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.

@@

To summarize, by transforming data to beamspace the complexity of all subsequent steps is potentially reduced. Calculation of the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculation of weights and the beamformer output are both $O(N_b)$. The added complexity of SBS is $O(N_bL)$. To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.

@@

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
\hfil
\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
\label{fig:benchBS96}}}
\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig. \ref{fig:bench}. When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively. The large jumps in execution times are caused by adaptive kernel launch configurations, which depends on $M$, $L$ and $N_b$.}
\label{fig:benchBS}
\end{figure*}
In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency. Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line. The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007}, 
\begin{equation}\label{eq:rayleigh}
\delta{}\theta = \lambda/(2Mp),
\end{equation}
where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
To sample a $70\degree$ sector with $64$- and $96$-element arrays with $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.

The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory. This is a valid assumption, since all imaging scenarios presented in this paper are well below the 8.0 GB/s limit of PCI Express 2.0. What is included are all calculations, memory transfers from global to shared GPU memory, and shared memory access.

Benchmarks of the ES-Capon beamformer are presented in Fig. \ref{fig:bench}. For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$. For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 8.3 frames per second (FPS). For $M=96$, $L=48$ and $K=2$ the processing throughput drops below $0.5$ FPS.

In Fig. \ref{fig:benchBS} we present similar benchmarks for the BS-Capon beamformer. The figure only presents results with use of temporal averaging ($K=2$). For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 FPS (8 ms). For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 FPS (13 ms). In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using one receive line per transmit. Hence, below this line we are capable of performing real-time processing.

\section{Simulated phantom and \textit{in-vivo} cardiac images}\label{sec:images}

\begin{figure*}[!t]
\centerline{
\subfloat[]{ \includegraphics[width=2.2in]{gfx/das.png} \label{fig:phantomDAS} }
\hfil
\subfloat[]{ \includegraphics[width=2.2in]{gfx/capon.png} \label{fig:phantomMV} }
\hfil
\subfloat[]{ 
\includegraphics[width=2.2in]{gfx/capon_bs.png} \label{fig:phantomBS} 
\hfil
\includegraphics[width=0.35in]{gfx/colorBarPhantom2.png}
}
}
\caption{Simulated phantom containing one bright (20 dB) and one dark (-20 dB) circular area and two point targets (30 dB) in speckle (0 dB). The simulation was done using Field II with a 64-element, 3.4 MHz, and $\lambda/2$ pitch phased array. Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB. The image is the 15th frame in the attached videos. a) Delay-and-sum with uniform weights. b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 1}. c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 2}.}
\label{fig:phantom}
\end{figure*}

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=2.2in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
\hfil
\subfloat[]{\includegraphics[width=2.2in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
\hfil
\subfloat[]{
\includegraphics[width=2.2in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
\hfil
\includegraphics[width=0.4in]{gfx/colorBarInvivo2.png}
}
\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5 MHz, 64 elements phased array. Number of receive beams for the $70\degree$ sector is 98, and number of samples in range is 610. Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB. The image is the first frame in the attached videos. a) Delay-and-sum with uniform weights \multimedia{Media-Movie 3}. b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 4}. c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 5}.}
\label{fig:invivo}
\end{figure*}

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
\hfil
\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
}
\caption{Lateral slices taken from the images in Fig. \ref{fig:phantom} and Fig. \ref{fig:invivo}. a) From Fig. \ref{fig:phantom} at 93 mm constant depth, through the two bright point scatterers. b) From Fig. \ref{fig:invivo} at 96 mm range, through the tip of the mitral valve.}
\label{fig:slices}
\end{figure*}

To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig. \ref{fig:phantom}). The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth. Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth. The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon. The presented image is frame $15$ in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.

Fig. \ref{fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle. The image was acquired using a 64-element and $p=0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive). The image is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 3, 4 and 5}. It is included to show that both ES-Capon and BS-Capon beamforming are feasible for \textit{in-vivo} cardiac imaging, not just in simulations.

To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig. \ref{fig:phantom} and \ref{fig:invivo} are included in Fig. \ref{fig:slices}. For the phantom, the cut goes through the two point targets at 90 mm constant radius, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant radius.

Both data sets have been processed with the following settings for the ES-Capon beamformer: $L = 32$, $K = 2$, and $d=0.1$. These settings have previously been shown to provide improved lateral resolution while maintaining DAS-like speckle \cite{Synnevag2007a}. For BS-Capon the following parameters have been used: $L=32$, $K=2$, $d=0.01$, and $N_b=3$.

When the simulated phantom was processed with the ES-Capon beamformer, modulation effects were observed when the two bright point scatterers moved across receive-beams. The artifact did not disappear until 16x lateral oversampling relative to (\ref{eq:rayleigh}) was used on transmit. The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}). Despite this, the same artifacts were not observed here. The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.
